Supplemental Material for ``Assessing United States county-level exposure for research on tropical cyclones and human health"

### 2019-11-07 writing/ehp_results.tex
Make more edits

We created and published this tropical cyclone exposure data as open-source
data, so they can be used by epidemiological
researchers~\parencite{hurricaneexposuredata}. These data include continuous
county-level measurements for the closest distance of each storm, cumulative
rainfall, and maximum sustained surface wind speed. These continuous metrics
can be used for classifying counties as exposed or unexposed, using thresholds
selected by the user, or can be used as continuous metrics. The dataset also
include binary data on flood and tornado events associated with each storm in
each county. To accompany these data, we published software tools to explore
and map the data and to integrate it with human health
datasets~\parencite{hurricaneexposure}.

We explored potential limitations in these data by comparing them with data
from other available sources.  For estimates of storm-associated rainfall, we
compared estimates from NLDAS-2 re-analysis precipitation data (used as the
rainfall estimates in the open-source data) with ground-based observations in
nine sample counties (Figure~\ref{fig:raincomparison}). Within these counties,
storm-related rainfall measurements were well-correlated between the two data
sources, with rank correlations (bottom right of each graph in
Figure~\ref{fig:raincomparison}) between 0.87 and 0.98. There was some evidence
that the reanalysis data may tend to underestimate rainfall totals in storms
with extremely high rainfall, based on a few heavy-rainfall storms in Harris
County, TX, Mobile County, AL, Charleston County, SC, and Wake County, NC
(Figure~\ref{fig:raincomparison}). However, even in these cases, it was rare
for a storm to be classified differently (under the precipitation threshold
of 75~\si{\milli\metre} we used for binary exposure classification for later
analysis) based on the source of precipitation data---horizontal and vertical
lines in each small plot in Figure~\ref{fig:raincomparison} show the threshold
of 75~\si{\milli\metre}, so storms in the lower left and upper right quadrants
would be classified the same (``exposed'' or ``unexposed'') regardless of the
precipitation data source, while storms in the upper left and lower right
quadrants would be classified differently. Such cases are rare.

For maximum sustained surface wind estimates, we found that the modeled
estimates, which we included as the primary wind exposure metric in the open
source data and used for further analysis in this study, generally agreed well
with estimates based on the wind radii reported in \ac{HURDAT2}. For most
storms, $\ge$95\% of counties were assigned the same category of wind speed
($<$34~\si{\knot}; 34\,--\,49.9~\si{\knot}; 50\,--\,63.9~\si{\knot};
$\ge$64~\si{\knot}) by both data sources (Figure~\ref{fig:windcomparison}).
Disagreement was limited to large storms that were characterized by high winds
well inland, like Hurricanes Sandy in 2012 and Ike in 2008. For these two
storms, the modeled wind speed values somewhat overestimated the extent of the
storm's most severe winds at landfall but then underestimated the extent of
34\,--\,49.9 \si{\knot} storm winds further inland (Figure~S2). For
epidemiology researchers who would like to conduct sensitivity analysis
between these two sets of wind data, we have included estimates from the
\ac{HURDAT2} wind radii as a secondary measure of wind speeds in the
open-source dataset \parencite{hurricaneexposuredata}.

For flood data, we compared flood status, based on event listings in the NOAA
Storm Events database and included in the open-source data, to streamflow
measurements at \ac{US} Geological Survey gages within nine sample counties
(Figure~\ref{fig:floodcomparison}). Across the sampled counties, streamgage
data generally indicated higher discharge during periods identified as flood
events based on the NOAA Storm Events database, particularly in counties with
more reporting streamgages (e.g., Harris County, TX, with~10 reporting stream
gages; Figure~\ref{fig:floodcomparison}). There were some cases, however, where
the two flooding data sources were somewhat inconsistent.  For example, there
were one or two tropical cyclones in several of the counties (Mobile County,
AL, Escambia County, FL, and Montgomery County, PA) with associated flood event
listings but for which the total discharge across county streamflow gages was
relatively low. For storms without a flood listing for the county, in most
cases the total streamflow discharge in the county was relatively low, and, in
all but one of these cases all streamgage flows were below the flooding
threshold. The exception was for one storm in Fulton County, GA (Hurricane Ida
in 2009).

@@

We also measured the average number of counties exposed per storm to each
hazard (limiting this analysis to tropical cyclones that exposed at least one
county to that hazard; Table~\ref{tab:exposuresummaries}).  If a tropical
cyclone exposed any counties to tornadoes, it was typically only a few (median
counties exposed per storm:~7). When a storm exposed any counties to other
tropical cyclone exposures, the swath of exposed counties tended to be much
larger: a median of~25 counties exposed for high winds per storm,~26 for
flooding, and~68 for extreme rainfall.  For every hazard except tropical
cyclone tornadoes, we identified at least one tropical cyclone that exposed
over~250 counties (Table~\ref{tab:exposuresummaries}).  However, the
largest-extent tropical cyclone varied across hazards: Frances in~2004 exposed
the most counties based on rain, Ike in~2008 based on wind, and Ivan in 2004
based on flooding and tornadoes (Table~\ref{tab:exposuresummaries}).

When we calculated and mapped the average number of exposures per decade in
each county for single-hazard exposures (Figure~\ref{fig:averageexposure}),
strong geographical patterns were clear. Tropical cyclone wind exposure had a
strong coastal pattern, with almost all exposures in counties within about~200
\si{\kilo\metre} (124~mi) of the coastline. While tropical cyclone rain
exposure was also more frequent in coastal areas compared to inland areas,
there were also inland exposures in counties that were never to extreme storm
winds. Rain and, to some extent, flood exposures were characterized by a
pattern defined by the Appalachian Mountains, with fewer exposures west of the
mountain range than to the east. Almost all tornado-based exposures were in
coastal states, with many in Florida, and almost none north of Maryland.
Flood-based exposures were frequently in the Mid-Atlantic region, with a
pattern that skewed north compared with other exposures.  Patterns were similar
when analysis was restricted to years with exposure data for all four hazards
available (1996\,--\,2011; Figure~S3).

@@

Next, we measured the agreement between exposure classifications for each pair
of hazards, as well as between each hazard and exposure assessment based on
distance from the storm's track as a proxy for exposure to storm hazards. We
found that these exposure classifications typically did not strongly agree
between pairs of metrics. The set of counties identified as ``exposed'' based
on one metric often overlap little with the set identified as ``exposed'' by
another metric, as storms frequently brought different hazards to different 
locations.

Figure~\ref{fig:ivanexposure} shows shows an example: Hurricane Ivan in 2004.
For the distance-based metric, the counties assessed as exposed follow the
tropical cyclone's track. For the wind-based metric,
only counties near the tropical cyclone's first landfall were assessed as
exposed. For rain- and flood-based metrics, however, exposure extended to the
left of the track, including counties as far north as New York and Connecticut,
while for the tornado metric, exposed counties tended to be to the right of the
track and included several counties in central North Carolina, South Carolina,
and Georgia that were not identified as exposed to Ivan based on any other
metric. Figure~S4 provides similar maps for four other example tropical
cyclones (selected because they exposed many \ac{US}  counties based on at
least one metric).

We drew similar conclusions when we investigated all~46 tropical cyclones
between~1996 and~2011 (the period for which data for all five metrics were
available) for which~100 or more counties were exposed based on at least one
metric (Figure~\ref{fig:jaccard}). In this figure, each row provides results
for one tropical cyclone, and each box in that row shows the measured Jaccard
coefficient for a pair of metrics. For most tropical cyclones,
distance-based county-level exposure assessment was, at best, in moderate
agreement with exposure assessments based on the four hazard-based metrics. For
almost all tropical cyclones, Jaccard index values between distance and any of
the four hazard-based metrics were~$\le0.6$ (i.e., out of the counties assessed
as exposed by at least one of the two metrics in the pair,~60\si{\percent} or
fewer were assessed the same under both metrics in the pair). For many tropical
cyclones, the Jaccard index values were~$\le0.4$, indicating poor agreement.
Among the four hazard-based metrics, the tornado-based metric had
universally poor agreement with other metrics in county-level classification
across the tropical cyclones considered.

There were, however, exceptions\,---\,tropical cyclones in which exposure
assessment agreed well across several of the metrics considered.  For Floyd
in~1999 (Figure~S4) and Irene in~2011, for example, county-level classification
agreed moderately to well for all pairs of exposure metrics except those
including the tornado-based metric (Figure~\ref{fig:jaccard}). For another set
of tropical cyclones (e.g., Ernesto in~2006, and Bertha in~1996, Isabel
in~2003), there was moderate to good agreement for all pairwise combinations of
distance, rain, and wind, but poor agreement for all other combinations of
metrics, while for another set of storms (e.g., Matthew in~2004 and Katrina
in~2005), there was moderate to good agreement between distance and rain.
However, these sets of tropical cyclones represented the minority of tropical
cyclones considered; for most tropical cyclones, there was low overlap in the
counties determined to be exposed to the tropical cyclones between most
pairings of metrics.

\subsection*{Case study}

We conducted a case study comparing estimates of physical exposure to
tropical cyclones that were measured using each exposure metric.
Figure~\ref{fig:topelecdependexposure} shows the study counties with the
highest expected rates of physical exposure to tropical cyclones among
electricity-dependent Medicare beneficiaries based on each metric. A few
counties (Miami-Dade County, FL; Harris County, TX) were ranked in the top
three under almost all exposure metrics
(Figure~\ref{fig:topelecdependexposure}). However, there were key differences
across exposure metrics. For example, most of the top counties based on wind-
and tornado-based metrics were in Florida, while over half of the top counties
based on the flood-based metric were in Mid-Atlantic states (PA, NY).

>>>>

\textit{We created and published this tropical cyclone exposure data as open-source
data, so they can be used by epidemiological
researchers~\parencite{hurricaneexposuredata}. These data include continuous
county-level measurements for the closest distance of each storm, cumulative
rainfall, and maximum sustained surface wind speed. These continuous metrics
can be used for classifying counties as exposed or unexposed, using thresholds
selected by the user, or can be used as continuous metrics. The dataset also
include binary data on flood and tornado events associated with each storm in
each county. To accompany these data, we published software tools to explore
and map the data and to integrate it with human health
datasets~\parencite{hurricaneexposure}.}

We explored potential limitations in these data by comparing them with data
from other available sources.  For estimates of storm-associated rainfall, we
compared estimates used in the open-source data with ground-based observations
in nine sample counties (Figure~\ref{fig:raincomparison}). Within these
counties, storm-related rainfall measurements were well-correlated between the
two data sources, with rank correlations (bottom right of each graph in
Figure~\ref{fig:raincomparison}) between 0.87 and 0.98. There was some evidence
that the reanalysis data included in the open-source dataset may tend to
underestimate rainfall totals in storms with extremely high rainfall, based on
a few heavy-rainfall storms in Harris County, TX, Mobile County, AL, Charleston
County, SC, and Wake County, NC (Figure~\ref{fig:raincomparison}). However,
even in these cases, it was rare for a storm to be classified differently
(under the precipitation threshold of 75~\si{\milli\metre} we used for binary
exposure classification for later analysis) based on the source of
precipitation data---horizontal and vertical lines in each small plot in
Figure~\ref{fig:raincomparison} show the threshold of 75~\si{\milli\metre}, so
storms in the lower left and upper right quadrants would be classified the same
(``exposed'' or ``unexposed'') regardless of the precipitation data source,
while storms in the upper left and lower right quadrants would be classified
differently. Such cases are rare.

For maximum sustained surface wind estimates, we found that the primary wind
exposure metric in the open source data generally agreed well with the wind
radii reported in \ac{HURDAT2}. For most storms, $\ge$95\% of counties were
assigned the same category of wind speed ($<$34~\si{\knot};
34\,--\,49.9~\si{\knot}; 50\,--\,63.9~\si{\knot}; $\ge$64~\si{\knot}) by both
data sources (Figure~\ref{fig:windcomparison}).  Disagreement was limited to a
few storms (e.g., Hurricanes Sandy in 2012 and Ike in 2008). For these two
storms, the modeled wind speed values used in the open-source data somewhat
overestimated the extent of the storm's most severe winds at landfall but then
underestimated the extent of 34\,--\,49.9 \si{\knot} storm winds further inland
(Figure~S2). For epidemiology researchers who would like to conduct sensitivity
analysis between these two sets of wind data, we have included estimates from
the \ac{HURDAT2} wind radii as a secondary measure of wind speeds in the
open-source dataset \parencite{hurricaneexposuredata}.

For flood data, we compared the flood status values included in the open-source
data to streamflow measurements at \ac{US} Geological Survey gages within nine
sample counties (Figure~\ref{fig:floodcomparison}). Across the sampled
counties, streamgage data generally indicated higher discharge during periods
identified as flood events based on the NOAA Storm Events database,
particularly in counties with more reporting streamgages (e.g., Harris County,
TX, with~10 reporting stream gages; Figure~\ref{fig:floodcomparison}). There
were some cases, however, where the two flooding data sources were somewhat
inconsistent.  For example, there were one or two tropical cyclones in several
of the counties (Mobile County, AL, Escambia County, FL, and Montgomery County,
PA) with associated flood event listings but for which the total discharge
across county streamflow gages was relatively low. For storms without a flood
listing for the county, in most cases the total streamflow discharge in the
county was relatively low, and, in all but one of these cases all streamgage
flows were below the flooding threshold. The exception was for one storm in
Fulton County, GA (Hurricane Ida in 2009).

@@

\textit{We also measured the average number of counties exposed per storm to each
hazard (limiting this analysis to tropical cyclones that exposed at least one
county to that hazard; Table~\ref{tab:exposuresummaries}).  If a tropical
cyclone exposed any counties to tornadoes, it was typically only a few (median
counties exposed per storm:~7). When a storm exposed any counties to other
tropical cyclone exposures, the swath of exposed counties tended to be much
larger: a median of~25 counties exposed for high winds per storm,~26 for
flooding, and~68 for extreme rainfall.  For every hazard except tropical
cyclone tornadoes, we identified at least one tropical cyclone that exposed
over~250 counties (Table~\ref{tab:exposuresummaries}).  However, the
largest-extent tropical cyclone varied across hazards: Frances in~2004 exposed
the most counties based on rain, Ike in~2008 based on wind, and Ivan in 2004
based on flooding and tornadoes (Table~\ref{tab:exposuresummaries}).}

When we calculated and mapped the average number of exposures per decade in
each county for single-hazard exposures (Figure~\ref{fig:averageexposure}),
strong geographical patterns were clear. Tropical cyclone wind exposure had a
strong coastal pattern, with almost all exposures in counties within about~200
\si{\kilo\metre} (124~mi) of the coastline. While tropical cyclone rain
exposure was also more frequent in coastal areas compared to inland areas,
there were also inland exposures in counties that were never classified as
exposed based on winds. Flood-based exposures were frequently in the
Mid-Atlantic region, with a pattern that skewed north compared with other
exposures.Rain and, to some extent, flood exposures were characterized by a
pattern defined by the Appalachian Mountains, with fewer exposures west of the
mountain range than to the east. Almost all tornado-based exposures were in
coastal states, with many in Florida, and almost none north of Maryland.
Patterns were similar when analysis was restricted to years with exposure data
for all four hazards available (1996\,--\,2011; Figure~S3).

@@

Next, we measured the agreement between exposure classifications for each pair
of hazards, as well as between each hazard and exposure assessment based on
distance from the storm's track. We found that these exposure classifications
typically did not strongly agree between pairs of metrics---the set of counties
identified as ``exposed'' based on one metric often overlapped little with the set
identified as ``exposed'' by another metric, as storms frequently brought
different hazards to different locations.

Figure~\ref{fig:ivanexposure} shows as an example Hurricane Ivan in 2004.
For the distance-based metric, the counties assessed as exposed follow the
tropical cyclone's track. For the wind-based metric,
only counties near the tropical cyclone's first landfall were assessed as
exposed. For rain- and flood-based metrics, however, exposure extended to the
left of the track, including counties as far north as New York and Connecticut,
while for the tornado metric, exposed counties tended to be to the right of the
track and included several counties in central North Carolina, South Carolina,
and Georgia that were not identified as exposed to Ivan based on any other
metric. Figure~S4 provides similar maps for four other example tropical
cyclones (selected because they exposed many \ac{US}  counties based on at
least one metric).

We drew similar conclusions when we investigated all~46 tropical cyclones
between~1996 and~2011 (the period for which data for all five metrics were
available) for which~100 or more counties were exposed based on at least one
metric (Figure~\ref{fig:jaccard}). In this figure, each row provides results
for one tropical cyclone, and each box in that row shows the measured Jaccard
coefficient for a pair of metrics.  For almost all tropical cyclones, Jaccard
index values between distance and any of the four hazard-based metrics
were~$\le0.6$ (i.e., out of the counties assessed as exposed by at least one of
the two metrics in the pair,~60\si{\percent} or fewer were assessed the same
under both metrics in the pair). For many tropical cyclones, the Jaccard index
values were~$\le0.5$, indicating at best moderate agreement.  Among the four
hazard-based metrics, the tornado-based metric had universally poor agreement
with other metrics in county-level classification across the tropical cyclones
considered.  Agreement between other pairs of metrics were better, but still
typically at best moderate (Jaccard index values usually $\le0.5$).

There were, however, exceptions\,---\,tropical cyclones in which exposure
assessment agreed well across several of the metrics considered.  For Floyd
in~1999 (Figure~S4) and Irene in~2011, for example, county-level classification
agreed moderately to well for all pairs of exposure metrics except those
involving the tornado-based metric (Figure~\ref{fig:jaccard}). For another set
of tropical cyclones (e.g., Ernesto in~2006, and Bertha in~1996, Isabel
in~2003), there was moderate to good agreement for all pairwise combinations of
distance, rain, and wind, but poor agreement for other combinations of
metrics, while for another set of storms (e.g., Matthew in~2004 and Katrina
in~2005), there was moderate to good agreement between distance and rain.
However, these sets of tropical cyclones represented the minority of tropical
cyclones considered; for most tropical cyclones, there was low overlap in the
counties determined to be exposed to the tropical cyclones between most
pairings of metrics.
